Non-word repetition in Yélî Dnye: A small-scale study

### 2019-10-30 manuscript.Rmd
results in progress

- What is the overall repetition accuracy (whole word, phoneme based, distance)?
- How does this change as a function of item complexity (number of syllablex, sound complexity)?
- How frequent are errors that result in real words? Is that a function of item complexity? What are some common patterns of phonological errors?
- Is individual variation explainable by child age, sex, birth order, monolingual status, and/or parental education?

@@

@ac ADD TABLE SUMMARIZING PREVIOUS WORK

- age differences
- length effects

@@

```{r cars}
# read.csv("NWR-test-log_FINAL.csv",header=T,sep=";")->fulldemo
# fulldemo=fulldemo[,-2] #remove children's names
# fulldemo$dob=as.Date(as.character(fulldemo$DOB.ISO),format="%Y%m%d")
# fulldemo$test=as.Date(as.character(fulldemo$Date),format="%Y%m%d")
# fulldemo$age.exact=as.numeric(difftime(fulldemo$test,fulldemo$dob,unit="weeks"))/52.25
# fulldemo$Age=as.numeric(as.character(fulldemo$Age))
# plot(fulldemo$Age~fulldemo$age.exact)
# write.csv(fulldemo,"nwr-demo.csv")

@@

@@

```{r}
read.table("log2.txt",header=T,sep="\t")->x
x$rep=gsub(".wav","",gsub(".*_","",x$out),fixed=T)
x[order(x$out),]->x
x$cor.online=1
for(i in 1:dim(x)[1]) if(x$rep[i]>1) x$cor.online[i-1]<-0

subset(x,rep=="1")->x
aggregate(x$cor.online,by=list(x$child),mean)->onl
colnames(onl)<-c("ID","score.online")

print(onl)

aggregate(x$corr,by=list(x$child),mean)->ms
colnames(ms)<-c("ID","score")
merge(ms,onl)->ms
merge(ms,demo)->scores
scores$Age=as.numeric(as.character(scores$Age))
scores$Age[is.na(scores$Age)]<-15
plot(scores$score*100~scores$Age,pch=20,xlab="Age (years)",ylab="% Correct")
abline(lm(scores$score*100~scores$Age))
plot(scores$score~scores$monolingual)
points(scores$score~as.numeric(scores$monolingual),pch=20)

plot(ms$score.online~ms$score)
```

@@

- What is the overall repetition accuracy (whole word, phoneme based, distance)?
- How does this change as a function of item complexity (number of syllablex, sound complexity)?
- How frequent are errors that result in real words? Is that a function of item complexity?
- Is individual variation explainable by child age, sex, birth order, monolingual status, and/or parental education?

>>>>

- What is the overall repetition accuracy (whole word, phoneme based, distance)?
- How does this change as a function of item complexity (number of syllables, sound complexity)?
- How frequent are errors that result in real words? Is that a function of item complexity? What are some common patterns of phonological errors?
- Is individual variation explainable by child age, sex, birth order, monolingual status, and/or parental education?

@@

```{r prevlit-fig, echo = FALSE, ig.pos = "T!", fig.width=8, fig.height=5.5, fig.cap = "Proportion of non-words correctly repeated as a function of age (in years), study (first author and year regardless of the number of authors), and, when available, shorter versus longer non-words."}
read.csv("prevlit.csv")->pl
#duplicate Piazzalunga logging short and long items separately
pl$Score[pl$Study=="Piazzalunga 2018"]<-pl$Short[pl$Study=="Piazzalunga 2018"]
pl$Study<-as.character(pl$Study)
pl$Study[pl$Study=="Piazzalunga 2018"]<-"Piazzalunga 2018 1-2 syll"
dim(pl)[1]->beg
pl=rbind(pl,pl[pl$Study=="Piazzalunga 2018 1-2 syll",])
pl[(beg+1):dim(pl)[1],"Score"]<-pl[(beg+1):dim(pl)[1],"Long"]
pl[(beg+1):dim(pl)[1],"Study"]<-"Piazzalunga 2018 3-4 syll"
pl$Study=paste(pl$Language,pl$Study)
library(lattice)
xyplot(Score~Age,groups=Study,data=pl,type=c('l','p'),auto.key = T ,pch=20)

@@

```{r demo}
# read.csv("NWR-test-log_FINAL.csv",header=T,sep=";")->fulldemo
# fulldemo=fulldemo[,-2] #remove children's names
# fulldemo$dob=as.Date(as.character(fulldemo$DOB.ISO),format="%Y%m%d")
# fulldemo$test=as.Date(as.character(fulldemo$Date),format="%Y%m%d")
# fulldemo$age.exact=as.numeric(difftime(fulldemo$test,fulldemo$dob,unit="weeks"))/52.25
# fulldemo$Age=as.numeric(as.character(fulldemo$Age))
# plot(fulldemo$Age~fulldemo$age.exact)
# write.csv(fulldemo,"nwr-demo.csv")

@@

```{r wrangling}
#@mc look here - we probably need to read the table in a special way to get all the accents etc. right
read.table("NWR-transcription.txt",header=T,enc="UTF-16le")->trnsc
read.csv("NWR-transcription.csv",sep=";",enc="UTF-8")->trnsc
trnsc$tokid=paste(trnsc$item,trnsc$token)
#task 4 are probably the items with no target; there are 1488 lines of paired data

read.table("final_order.txt",header=T)->crp
crp$tokid=paste(crp$target,crp$nb)
npairs=dim(crp)[1]

trnsc=merge(trnsc,crp,by.x="tokid",by.y="tokid",all=T)

# sort by file name which indicates presentation order, so that we can get the first attempt
trnsc[order(trnsc$outfile),]->trnsc

#add repetition number
trnsc$rep=gsub(".*_","",gsub(".wav","",trnsc$outfile))

#add coding of correctness online/on the fly of first presentation as a function whether there are subsequent reps of the same target
trnsc$cor.online=1
for(i in 1:npairs) if(trnsc$rep[i]>1) trnsc$cor.online[i-1]<-0  #if the item is repeated, then code the previous one as being wrong
trnsc$cor.online[(npairs+1):dim(trnsc)[1]]<-NA

#DATA REMOVAL *** ATTENTION *** 
#REMOVING ALL THE "YI"
trnsc[trnsc$target!="yi",]->trnsc
#REMOVE ALL TASK 4
trnsc[!is.na(trnsc$target),]->trnsc

```{r phonoana}

#convert target coding to orthography
orthotarget=matrix( #list correspondances always by pairs: orthography then file name; for longer items orthography=filename
  c("dpa","dpa",
    "dp:a","dp_a",
    "dpâ","dpa2",
    "dpéé","dpee",
    "dpê","dpe",
    "dpi","dpi",
    "dpu","dpu",
    "kp:ââ","kp_aa",
    "kpu","kpu",
    "tp:a","tp_a",
    "tpâ","tpa",
    "tpê","tpe",
    "gh:ââ","gh_aa",
    "ghuu","ghuu",
    "lva","lva",
    "lv:ê","lv_e",
    "lvi","lvi",
    "t:êê","t_ee"
  ),#last item above should not have a comma
  ncol = 2,byrow = T)

trnsc$target.ortho=as.character(trnsc$target) #to start with, put them all in
for(i in 1:dim(orthotarget)[1]) trnsc$target.ortho[trnsc$target.ortho==orthotarget[i,2]] <- orthotarget[i,1] #then correct the monosyllables

## TODO
# make all vowels in mispronunciation & target short (so we don't penalize length errors)
# if correct=0 but orthotarget=mispronunciation (probably due to length errors) then change correct to 1

# create phonological correspondance matrix
# NOTE!! NON INTUITIVE MAPPING!!!! DO NOT READ THE OUTPUT BELIEVING IT IS PSEUDO IPA!!!
correspondances=matrix( #list correspondances always by pairs, orthography then phonology
  c(
    #nasal long to unichar
    ":ii","ì",
    ":ee","è",
    ":aa","à",
    ":êê","3",
    ":ââ","6",
    ":uu","ù",
    ":oo","Ò",
    #long to unichar
    "ii","ï",
    "ee","ë",
    "éé","Ë",
    "áá","Ä",
    "aa","ä",
    "îî","Ï",
    "êê","2",
    "ââ","5",
    "uu","ü",
    "óó","ö",
    "oo","Ö",
    #short nasal
    ":i","1",
    ":e","4",
    ":a","ã",
    ":ê","6",
    ":â","7",
    ":u","8",
    ":o","9",
    #consonants, veeeery dirty!
    "'n","ń",
    "tp",":",
    "dp","/",
    "kp","P",
    "ngm","G",
    "ńm","M",
    "ng","N",
    "mb","b",
    "nd","D",
    "nt","T",
    "nk","K",
    "dn","?",
    "kn","!",
    "vy","V",
    "mt","C",
    "md","J",
    "mg","X",
    "dm","S",
    "km","Z",
    "gh","H",
    "lv","L"
  ),#last item above should not have a comma
  ncol = 2,byrow = T)

trnsc$mp_uni=trnsc$mispronunciation #start from ortho
trnsc$target_uni=trnsc$target.ortho #start from ortho

for(i in 1:dim(correspondances)[1]) { #transform into unicharacter
  trnsc$mp_uni=gsub(correspondances[i,1],correspondances[i,2],trnsc$mp_uni)
  trnsc$target_uni=gsub(correspondances[i,1],correspondances[i,2],trnsc$target_uni)
}

#log non-fluent speech
trnsc$nonfluent<-NA
trnsc$nonfluent[grep("-",trnsc$mp_uni,fixed=T)]<-1

#trnsc$additions=
```

```{r}
#create a subset with just the first presentations -- this is what is going to be used for the quantitative analyses
firstp=subset(trnsc,rep==1 & !is.na(rep))
```

@@

```{r iniana, eval=FALSE}
#initial analyses with just 12 participants
read.table("log2.txt",header=T,sep="\t")->x
x$rep=gsub(".wav","",gsub(".*_","",x$out),fixed=T)
x[order(x$out),]->x
x$cor.online=1
for(i in 1:dim(x)[1]) if(x$rep[i]>1) x$cor.online[i-1]<-0  #if the item is repeated, then code the previous one as being wrong

subset(x,rep=="1")->x  #get only the first attempt for each child
aggregate(x$cor.online,by=list(x$child),mean)->onl  #get means on first attempt
colnames(onl)<-c("ID","score.online")

#print(onl) #online/on the fly coding, there are only 12 kids

aggregate(x$corr,by=list(x$child),mean)->ms  #means
colnames(ms)<-c("ID","score")
merge(ms,onl)->ms
merge(ms,demo)->scores
scores$Age=as.numeric(as.character(scores$Age))
scores$Age[is.na(scores$Age)]<-15

# plot(scores$score*100~scores$Age,pch=20,xlab="Age (years)",ylab="% Correct")
# abline(lm(scores$score*100~scores$Age))
# plot(scores$score~scores$monolingual)
# points(scores$score~as.numeric(scores$monolingual),pch=20)

#plot(ms$score.online~ms$score)
```

@@

- What is the overall repetition accuracy (whole word, phoneme based, distance)?
- How does this change as a function of item complexity (number of syllables, sound complexity)?
- How frequent are errors that result in real words? Is that a function of item complexity?
- Is individual variation explainable by child age, sex, birth order, monolingual status, and/or parental education?
